What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-05-07 android-stacks.tex
mprovements on section III.

%\subsection{Error handling in programming languages}

@@

\subsection{Exception handling in Java} \label{sec:extypes}

@@

When an exception is signaled it causes a transfer of control from the point
where the exception occurred to a point that can be specified by the programmer
(exception handler) - in Java it is represented by the try-catch block .

%and is
%responsible for performing any recovery action necessary to bring the software
%system back to a normal state and, whenever this is not possible, to log the
%exception and abort the system in an pre-defined safe way.

%In Java, as in many languages, 
The search for the handler to deal with a raised
exception occurs along the dynamic invocation chain. 
%This is claimed to increase
%the reusability, since the invoker of an operation can handle it in a wider
%context ~\cite{goodenough1975exception}. 
If no handler is found in the chain, the default handler associated to the 
Thread or ThreadGroup is invoked, which usually terminates the system 
and prints the exception stack trace.

@@

Figure~\ref{fig:wrapping} presents an example of a stack trace, on which multiple 
exception chaining is depicted. Exception Chaining (or exception wrapping) is  
common way of propagating exceptions  across components of different 
abstraction levels. It is implemented by special exception 
constructor which receives another exception as a parameter.

@@

%\todo{This section feels a bit disconnected.}
Several general guidelines have been proposed on how to use
exceptions~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,
bloch2008effective}. To guide our investigation, we compile general guidelines
related to exception types and the propagation of exception in Java programs
\footnote{We also compiled guidelines related to exception handling but they are
out of the scope of this paper, since they do not relate to information
available on stack traces.}.

\noindent\emph{Meaning of Exception Types}

\textbf{1) Use checked exceptions to represent recoverable
conditions}(~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The developer should use checked exceptions for conditions from which the caller
is expected to recover. By confronting the API user with a checked exception,
the API designer is telling the client to handle the exceptional condition. The
client can explicitly ignore the exception (swallowing, or converting it to
other type) at the expense of the program's robustness~\cite{gosling2000java}.

\textbf{2) Error represents an unrecoverable condition detected by the JVM which
should not he handled}~\cite{gosling2000java}. It results from failures detected
by the Java Virtual Machine which indicate resource deficiencies, invariant
failures or other conditions that make impossible the program to recover.

\noindent\emph{Exception Throwing}

\textbf{3) A method should throw exceptions that precisely define the
exceptional condition}~\cite{gosling2000java,bloch2008effective}. To do so,
developers should either try to reuse the exception types already defined in the
JVM or they should create a specific exception. Throwing general types such as a
pure Exception or a RuntimeException is hence considered a bad practice.

\textbf{4) Do not repeatedly re-throw exceptions, handle exceptions as close as
possible to the problem}~\cite{wirfs2006toward}. Programs that frequently throw
exceptions lose in performance ~\cite{wirfs2006toward,gosling2000java}.
Moreover, close to the place that the exception was throw the caller usually has
enough contextual information to perform  corrective actions. When the exception
is propagated far away from the method where it was signaled (i.e. further away
from the problem) it can be difficult to take meaningful recovery actions.

\noindent\emph{Exception Documentation}

\textbf{5) Document all exceptions thrown by a
method}~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The exceptions thrown by a method are an important part of methods interface,
and is required to use the method properly. The checked exceptions are already
part of the  methods signature, and the method caller is aware of the checked
exceptions being thrown by it. According to~\cite{bloch2008effective}, it is
also wise to document the explicitly signaled runtime exceptions\footnote{This
excludes the implicitly signaled JVM runtime exceptions related to coding
errors} as carefully as checked exceptions. Doing so, the clients of a method
will be aware of the exceptions the method can throw. If the developer fails to
do follow this practice (specially when developing library code) it will be
difficult or even impossible for the caller to make effective use of such method
~\cite{wirfs2006toward, bloch2008effective}.

These best practices guided our exploratory study described in the next
sections. Based on information extracted from stack traces (complemented by
bytecode and source-code analysis), we investigate whether stack characteristics
can reveal whether these practices have been obeyed. Next section discusses the
study procedure and the practices' misuses that emerged from this study.

@@

Prevalence or prevalence proportion, a term often used in epidemiology,
represents the proportion of a population found to have a particular condition
(e.g. a disease) over the total number of people studied. In our work, we used
this term to evaluate a condition (i.e., presence of a stack trace) over issues
reported on GitHub repositories.

@@

To answer this question, we mined from every extracted stack trace the top level
exception being signaled, its signaler (the first method signature frame after
the exception), the root exception (the exception which initiated the stack
trace, defined on the bottom of the stack), and the root signaler. The
intermediate elements that compose a stack trace as shown in
Figure~\ref{fig:wrapping} (i.e., the intermediate causes and its execution
stacks) were not be considered in the analysis.
%~\todo{Move this to the previous section? IMO, It is better to keep all facts
%about how we did the research in one place}

@@

\subsection{Sizes of Stack traces}

Concerning the size of the stacks (i.e. the number of frames composing them), we
could observe that  sometimes it exceeded 100 frames (158 stack frames).
Manually inspecting the corresponding execution trace was caused by one of the
two reasons: recursive calls or exceptions signaled by a method defined deep
down in a reused framework; exceptions successively wrapped. Irrespective of the
reason, in exception flows involving many methods, it is practically impossible
to handle an exception in a way other than exiting the application, as at the
place the exception occurs there is hardly enough contextual information to
perform recovery actions.

\begin{figure}
  \begin{center}
    \includegraphics[scale=0.5]{stack-size-wrappings-hist.pdf}
  \end{center}
  \caption{Stack sizes (top) and wrappings (bottom) for Java (left) and Android (right) stack traces}
  \label{fig:sizewrapbhists}
\end{figure}

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories.

@@

As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions.

If we filter the exceptions implicitly signaled by JVM (due to coding erros) the number of runtime
exceptions thrown by third party code is XXX signaled by Android and YYY signaled by library code.

@@

>>>>

\subsection{Error handling in programming languages}

@@

\subsection{Exception Handling in Java} \label{sec:extypes}

@@

When an exception is signaled it causes a transfer of control from the point
where the exception occurred to a point that can be specified by the programmer
(exception handler) - in Java it is represented by the try-catch block .
The search for the handler to deal with a raised
exception occurs along the dynamic invocation chain. 
If no handler is found in the chain, the default handler associated to the 
Thread or ThreadGroup is invoked, which usually terminates the system 
and prints the exception stack trace. Figure~\ref{fig:wrapping} presents an 
example of a stack trace, on which multiple 
exception chaining is depicted. Exception Chaining (or exception wrapping) is  
common way of propagating exceptions  across components of different 
abstraction levels. It is implemented by special exception 
constructor which receives another exception as a parameter.

@@

@@

%\todo{This section feels a bit disconnected.}
Several general guidelines have been proposed on how to use Java
exceptions~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,
bloch2008effective}. 
% To guide our investigation, we compile the guidelines
A compiled list of the guidelines related to exception types and exception propagation in Java the following:
\footnote{We also compiled guidelines related to exception handling but they are
out of the scope of this paper, since they do not relate to information
available on stack traces.}:

%\noindent\emph{Meaning of Exception Types}

\textbf{I-Use checked exceptions to represent recoverable
conditions}(~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The developer should use checked exceptions for conditions from which the caller
is expected to recover. By confronting the API user with a checked exception,
the API designer is telling the client to handle the exceptional condition. The
client can explicitly ignore the exception (swallowing, or converting it to
other type) at the expense of the program's robustness~\cite{gosling2000java}.

\textbf{II-Error represents an unrecoverable condition detected by the JVM which
should not he handled}~\cite{gosling2000java}. It results from failures detected
by the Java Virtual Machine which indicate resource deficiencies, invariant
failures or other conditions that make impossible the program to recover.

%\noindent\emph{Exception Throwing}

\textbf{III-A method should throw exceptions that precisely define the
exceptional condition}~\cite{gosling2000java,bloch2008effective}. To do so,
developers should either try to reuse the exception types already defined in the
JVM or they should create a specific exception. Throwing general types such as a
pure Exception or a RuntimeException is hence considered a bad practice.

\textbf{IV-Do not repeatedly re-throw exceptions, handle exceptions as close as
possible to the problem}~\cite{wirfs2006toward}. 
%Programs that frequently throw exceptions lose in performance ~\cite{wirfs2006toward,gosling2000java}.
%Moreover, 
Close to the place that the exception was throw the caller usually has
enough contextual information to perform  corrective actions. When the exception
is propagated far away from the method where it was signaled (i.e. further away
from the problem) it can be difficult to take meaningful recovery actions.
todo{RSC:this practice was not well explored, can be removed if we need space.}

%\noindent\emph{Exception Documentation}
\textbf{V-Document all exceptions thrown by a
method}~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The exceptions thrown by a method are an important part of methods interface,
and is required to use the method properly. The checked exceptions are already
part of the  methods signature, and the method caller is aware of the checked
exceptions being thrown by it. According to~\cite{bloch2008effective}, it is
also wise to document the explicitly signaled runtime exceptions\footnote{This
excludes the implicitly signaled JVM runtime exceptions related to coding
errors} as carefully as checked exceptions. Doing so, the clients of a method
will be aware of the exceptions the method can throw. If the developer fails to
do follow this practice (specially when developing library code) it will be
difficult or even impossible for the caller to make effective use of such method
~\cite{wirfs2006toward, bloch2008effective}.

These best practices guided our exploratory study described next
. Based on information extracted from stack traces (complemented by
bytecode and source-code analysis), we investigate whether stack characteristics
can reveal whether these practices have been obeyed. 
%Next section discusses the
%study procedure and the practices' misuses that emerged from this study.

@@

Prevalence or prevalence proportion, a term often used in epidemiology,
represents the proportion of a population found to have a particular condition
(e.g. a disease) over the total number of the studied population. In our work, we used
this term to evaluate a condition (i.e., presence of a stack trace) over issues
reported on GitHub repositories.

@@

%To answer this question, we mined from every extracted stack trace the top level
%exception being signaled, its signaler (the first method signature frame after
%the exception), the root exception (the exception which initiated the stack
%trace, defined on the bottom of the stack), and the root signaler. The
%intermediate elements that compose a stack trace as shown in
%Figure~\ref{fig:wrapping} (i.e., the intermediate causes and its execution
%stacks) were not be considered in the analysis.
%~\todo{Move this to the previous section? IMO, It is better to keep all facts
%about how we did the research in one place
%RSC: I WAS ABOUT TO MOOVE BUT THERE WAS ALREADY A SIMILAR TEXT ON 
%THE EXCEPTION DISTEILLER}

@@

%\subsection{Sizes of Stack traces}
%Concerning the size of the stacks (i.e. the number of frames composing them), we
%could observe that  sometimes it exceeded 100 frames (158 stack frames).
%Manually inspecting the corresponding execution trace was caused by one of the
%two reasons: recursive calls or exceptions signaled by a method defined deep
%down in a reused framework; exceptions successively wrapped. Irrespective of the
%reason, in exception flows involving many methods, it is practically impossible
%to handle an exception in a way other than exiting the application, as at the
%place the exception occurs there is hardly enough contextual information to
%perform recovery actions.

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. If we filter the exceptions implicitly signaled by 
JVM (due to coding erros) the number of runtime exceptions thrown by third party code is XXX 
signaled by Android and YYY signaled by library code.

@@

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=0.5]{only-wrap-histogram.pdf}
  \end{center}
  \caption{Stack sizes (top) and wrappings (bottom) for Java (left) and Android (right) stack traces}
  \label{fig:sizewrapbhists}
\end{figure}
